---
title: "2007-2017 OASAS Admissions Analytics"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(devtools)
library("data.world")
library(paletteer)
library(MetBrewer)
```

```{r}
# Documentation from data.world recommends package installation from GitHub
devtools::install_github("datadotworld/data.world-r", build_vignettes = TRUE)

# Load the requisite API token obtained from data.world advanced settings
token <- readLines('api')
saved_cfg <- data.world::save_config(token)
data.world::set_config(saved_cfg)

sql_stmt <- data.world::qry_sql("SELECT * FROM 
                                chemical_dependence_treatment_program_admissions_beginning_2007_1")
admissions_df <- data.world::query(
  sql_stmt, "https://data.world/data-ny-gov/ngbt-9rwf")
```

---

## Data Understanding

* Initial look at dataframe, check data types

```{r}
str(admissions_df)
```

* Admissions trends grouped by each variable

```{r}
ggplot(admissions_df, aes(x = admissions)) +
  geom_histogram() + 
  labs(title = "Distribution of \"Admission\" Observations",
       x = "Admissions",
       y = "Count") +
  scale_x_log10() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5, margin = margin(t = 5))
  )

program_category_summary <- admissions_df %>% 
  group_by(program_category) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  ) 
print(program_category_summary)

service_type_summary <- admissions_df %>% 
  group_by(service_type) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
print(service_type_summary)

age_group_summary <- admissions_df %>% 
  group_by(age_group) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
print(age_group_summary)

primary_substance_group_summary <- admissions_df %>% 
  group_by(primary_substance_group) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
print(primary_substance_group_summary)
```

* Factoring variables and checking for outliers

```{r}
admissions_df <- admissions_df %>% 
  mutate(
    program_category = as.factor(program_category),
    service_type = as.factor(service_type),
    age_group = as.factor(age_group),
    primary_substance_group = as.factor(primary_substance_group)
  )

# Manual outlier identification
admissions_outliers <- admissions_df %>%
  mutate(
    mean_admissions = mean(admissions, na.rm = TRUE),
    sd_admissions = sd(admissions, na.rm = TRUE)
  ) %>%
  # Relative to the mean, any values on the lower or 
  # upper bounds that are 3 times the standard deviation are filtered
  filter(admissions < mean_admissions - 3 * sd_admissions |
         admissions > mean_admissions + 3 * sd_admissions) %>%
  select(admissions)

admissions_outliers

# Remove outliers
rmv_admissions_outliers <- admissions_df %>%
  mutate(
    mean_admissions = mean(admissions, na.rm = TRUE),
    sd_admissions = sd(admissions, na.rm = TRUE)
  ) %>%
  filter(!(admissions < mean_admissions - 3 * sd_admissions |
           admissions > mean_admissions + 3 * sd_admissions))

# Note subtracted outliers from new dataframe
str(admissions_df)
str(rmv_admissions_outliers$admissions)
```

* It may not be significant that admissions numbers are higher than usual at any given point in time

---

## Data Preparation

* Data is structured relationally prior to analysis

```{r}
# External county code .csv from https://www.dot.ny.gov
county_codes <- read_csv("county_codes.csv") 
county_codes
```

```{r}
county <- admissions_df %>%
  select(county_of_program_location) %>%
  distinct() %>%
  # Join codes with corresponding counties
  left_join(county_codes, by = "county_of_program_location") %>%
  mutate(county_code = case_when(
      # Tagging counties with first two characters and "-NYC"
      county_of_program_location %in% c("Bronx", "Queens", "Kings") ~
        paste(str_to_upper(str_sub(county_of_program_location, 1, 2)),
              "-NYC", sep = ""),
      county_of_program_location == "New York" ~ "NYC",
      # Handle to not treat as NA value, changed code to NS instead
      county_of_program_location == "Nassau" ~ "NS",
      # Handle to not treat as NA value, manually assigned SL 
      county_of_program_location == "St Lawrence" ~ "SL",
      TRUE ~ county_code 
    )
  )
county
```

```{r}
# Define abbreviations
program_category_index <- c(
  "Crisis" = "C",
  "Inpatient" = "I",
  "Opioid Treatment Program" = "OTP",
  "Outpatient" = "O",
  "Residential" = "R"
)
# Add program_code column
admissions_df <- admissions_df %>%
  mutate(program_code = recode(program_category, 
       "Crisis" = "C",
       "Inpatient" = "I",
       "Opioid Treatment Program" = "OTP",
       "Outpatient" = "O",
       "Residential" = "R"))

# Add program_category tibble 
program_category <- admissions_df %>%
  distinct(program_category, .keep_all = TRUE) %>%
  select(program_code, program_category)

program_category
```

* program_category contains identifiers assigned to each program category, duplicates

```{r}
# Define index
primary_substance_group_index <- c(
  "Alcohol" = "A",
  "All Others" = "AO",
  "Cocaine incl Crack" = "CC",
  "Heroin" = "H",
  "Marijuana incl Hashish" = "MH",
  "Other Opioids" = "OO"
)

# Add substance_code column
admissions_df <- admissions_df %>%
  mutate(substance_code = recode(primary_substance_group, 
        "Alcohol" = "A", 
        "All Others" = "AO",
        "Cocaine incl Crack" = "CC",
        "Heroin" = "H",
        "Marijuana incl Hashish" = "MH",
        "Other Opioids" = "OO"))

# Create substance dataframe based on index
primary_substance_group <- admissions_df %>%
  distinct(primary_substance_group, .keep_all = TRUE) %>%
  select(substance_code, primary_substance_group)
 
primary_substance_group
```

```{r}
# Join county_code by county name
admissions_df <- admissions_df %>% 
  full_join(county, by = "county_of_program_location")

admissions_df <- admissions_df %>% 
  select(
    year,
    county_code,
    program_code,
    service_type,
    age_group,
    primary_substance_group,
    substance_code,
    admissions
  )

str(admissions_df)
```

```{r}
# aggregate() to sum the total admissions for every year
annual_admissions <- function() {
  # Store aggregated data in separate tibble
  total_admissions <- aggregate(admissions_df$admissions, 
            by = list(year = admissions_df$year), 
            sum) %>% 
    rename(total = x)
  # Maximum point
  max_point <- total_admissions[which.max(total_admissions$total), ]
  # Line graph
  ggplot(total_admissions, aes(year, total)) + 
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = 2007:2017) +
    scale_y_continuous(limits = c(270000, 320000)) +
    labs(
      title = "OASAS Annual Admissions",
      x = "Year",
      y = "Total Admissions") +
    theme_minimal() +
    theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold",
                              margin = margin(t = 5, r = 10, l = 10, b = 10)),
    axis.title.x = element_text(size = 12, margin = margin(t = 10, b = 5)),
    axis.title.y = element_text(size = 12, margin = margin(r = 10, l = 5)),
    margin = margin(r = 10)
    ) + 
    annotate("text",
             x = max_point$year,
             y = max_point$total,
             label = paste("Top Admissions:", max_point$total, " in ", max_point$year),
             vjust = -1)
}

annual_admissions()
```

* Admissions peak in 2009 with 311,722 patients
* A significant downward trend is observed from 2010 to 2013
* Admissions continue to gradually decline from 2013 onward

---

## Data Evaluation

### Admissions by County

```{r}
# Total number of admission in NYS
total_admissions = sum(admissions_df$admissions)

# Calculate percentage of admissions in each county 
admissions_by_county <- admissions_df %>% 
  select(county_code, admissions) %>%
  group_by(county_code) %>% 
  summarize(percentage = ((sum(admissions) / total_admissions * 100))) %>%
  arrange(desc(percentage))

admissions_by_county

top_admissions <- admissions_by_county %>% 
  slice_max(n = 5, percentage)

top_admissions
```

### Rehab Facility Identification

```{r}
rehab_df <- admissions_df %>%
  # Only show rehabilitation services
  filter(str_detect(service_type, regex("Rehab|Rehabilitation",ignore_case = TRUE))) %>%
  select(service_type, age_group, primary_substance_group, admissions)
rehab_df

top_substance_df <- rehab_df %>%
  # Only interested in these combinations
  group_by(service_type, age_group, primary_substance_group) %>%
  # Take count to show how many admissions exist for each substance in each age group
  summarize(substance_count = sum(admissions)) %>%
  # Limit to age group 
  group_by(age_group) %>%
  # Filter for the substances with the highest count
  filter(substance_count == max(substance_count)) %>%
  # Show relevant columns
  select(service_type, age_group, primary_substance_group, substance_count)

top_substance_df
```

* To identify the most prominent substance used in each age group, we first define a regular expression in a new dataframe that filters all services containing "Rehab" or "Rehabilitation" in the name
* top_substance_df is created to find the top substance per age group
  * The data is grouped by age_group and primary_substance_group because we are only
  interested in analyses in the context of these variables paired together

### Admissions by Substance and Age Group

```{r}
all_substance_df <- rehab_df %>%
  group_by(service_type, age_group, primary_substance_group) %>%
  # Take count to show how many admissions exist for each substance in each age group
  summarize(substance_count = sum(admissions)) %>%
  # Limit to age group 
  group_by(age_group) %>%
  # Filter for the substances with the highest count
  mutate(substance_count == max(substance_count)) %>%
  # Show relevant columns
  select(service_type, age_group, primary_substance_group, substance_count)
all_substance_df
```

---
\pagebreak
## Data Visualization

### Top Admissions by County

```{r}
ggplot(top_admissions, aes(reorder(county_code, -percentage), percentage)) +
  geom_col(fill = "darkgray") + 
  labs(x = "County",
       y= "Percentage (%)",
       title = "Top OASAS Admissions by County") + 
  scale_y_continuous(expand = c(0,0)) +
  scale_x_discrete(labels = c("NYC" = "New York City",
                              "QU-NYC" = "Queens",
                              "BR-NYC" = "Bronx",
                              "KI-NYC" = "Kings",
                              "SU" = "Suffolk")) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5,
                              margin = margin(t = 5, b = 10)),
    axis.title.x = element_text(size = 12, color = "black",
                                margin = margin(t = 10, b = 5)),
    axis.title.y = element_text(size = 12, color = "black",
                                margin = margin(r = 10, l = 5))
  )
```

### Substance Use by Age Group

```{r}
ggplot(all_substance_df, aes(factor(age_group, 
                                    levels = c("Under 18", "18 thru 24",
                                               "25 thru 34", "35 thru 44",
                                               "45 thru 54", "55 and Older")), 
                             substance_count,
                             fill = primary_substance_group)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Age Group",
    y = "Number of Admissions",
    title = "Substance Use by Age Group",
    caption = "Stacked bar showing substance use by age group from 2007-2017.",
    legend = "Substance"
  ) +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_paletteer_d("MetBrewer::Archambault",
    name = "Substance",
    labels = c(
      "All Others" = "Other",
      "Cocaine incl Crack" = "Cocaine (incl. Crack)",
      "Marijuana incl Hashish" = "Marijuana (incl. Hashish)"
    )
  ) +
  scale_x_discrete(
    labels = c(
      "Under 18" = "<18",
      "18 thru 24" = "18-24",
      "25 thru 34" = "25-34",
      "35 thru 44" = "35-44",
      "45 thru 54" = "45-54",
      "55 and Older" = "55+"
    )
  ) +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 12, color = "black",
                                margin = margin(t = 10, b = 5)),
    axis.title.y = element_text(size = 12, color = "black",
                                margin = margin(r = 10, l = 10)),
    axis.text = element_text(color = "black"),
    legend.title = element_text(size = 12, color = "black", hjust = 0.5),
    plot.title = element_text(size = 14, face = "bold",
                              margin = margin(b = 10, t = 10), hjust = 0.5),
    plot.caption = element_text(size = 11, hjust = 0.5,
                                margin = margin(b = 5, t = 5))
  )
```

### Service Use by Age Group

```{r}
ggplot(all_substance_df, aes(factor(age_group, levels = c("Under 18", "18 thru 24",
                                                          "25 thru 34", "35 thru 44",
                                                          "45 thru 54", "55 and Older")),
                             substance_count,
                             fill = service_type)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    x = "Age Group",
    y = "Number of Admissions",
    title = "Service Use by Age Group",
    caption = "Stacked bar showing service use by age group from 2007-2017."
  ) +
  scale_fill_paletteer_d("palettetown::seedot", 
    name = "Service Type",
    labels = c(
      "Inpatient Rehabilitation" = "Inpatient",
      "Outpatient Rehabilitation" = "Outpatient",
      "Rehab and Reintegration" = "Rehab & Reintegration",
      "Res Rehab for Youth" = "Youth Residential",
      "Residential Rehabilitation" = "Residential",
      "Specialized Services OP Rehab" = "Specialized Services OP",
      "Stabilization and Rehab" = "Stabilization & Rehab"
    )
  ) +
  scale_x_discrete(
    labels = c(
      "18 thru 24" = "18-24",
      "25 thru 34" = "25-34",
      "35 thru 44" = "35-44",
      "45 thru 54" = "45-54",
      "55 and Older" = "55+",
      "Under 18" = "<18"
    )
  ) +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 12, color = "black",
                                margin = margin(t = 10, b = 5)),
    axis.title.y = element_text(size = 12, color = "black",
                                margin = margin(r = 10, l = 10)),
    axis.text = element_text(color = "black"),
    legend.title = element_text(size = 12, color = "black", hjust = 0.5),
    plot.title = element_text(size = 14, face = "bold",
                              margin = margin(b = 10, t = 10), hjust = 0.5),
    plot.caption = element_text(size = 11, hjust = 0.5,
                                margin = margin(b = 5, t = 5))
  )
```

### Top Substance Use by Age Group

```{r}
ggplot(top_substance_df, aes(primary_substance_group, substance_count,
                             fill = age_group)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +
  scale_y_continuous(expand = c(0,0)) +
  labs(
    y = "Number of Admissions",
    title = "Top Substance by Age Group",
    caption = "Stacked bar showing the top substance used per age group from 2007-2017."
  ) +
  scale_x_discrete(
    name = "Substance",
    labels = c(
      "All Others" = "Other",
      "Cocaine incl Crack" = "Cocaine (Includes Crack)",
      "Marijuana incl Hashish" = "Marijuana\n(Includes Hashish)"
    )
  ) +
  scale_fill_paletteer_d("MetBrewer::Archambault",
    name = "Age Group",
    labels = c(
      "18 thru 24" = "18-24",
      "25 thru 34" = "25-34",
      "35 thru 44" = "35-44",
      "45 thru 54" = "45-54",
      "55 and Older" = "55+",
      "Under 18" = "<18"
    )
  ) +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 12, color = "black",
                                margin = margin(t = 10, b = 5)),
    axis.title.y = element_text(size = 12, color = "black",
                                margin = margin(r = 10, l = 10)),
    axis.text = element_text(color = "black"),
    legend.title = element_text(size = 12, color = "black", hjust = 0.5),
    plot.title = element_text(size = 14, face = "bold",
                              margin = margin(b = 10, t = 10), hjust = 0.5),
    plot.caption = element_text(size = 11, hjust = 0.5,
                                margin = margin(b = 5, t = 5))
  )
```
